Local PCA Shows How Population Structure Differs Along the Genome

### 2016-08-10 writeup/local_pca_paper.tex
edited after chucks comments

In this work we describe variation in population structure using local PCA,
where ``local'' refers to proximity along the genome.
A number of general methods for dimensionality reduction also use a strategy of ``local PCA''
\citep[e.g.,][]{manjon2013diffusion,kambhatla1997dimension,weingessel2000local,roweis2000nonlinear},
performing PCA not on the entire dataset but instead on subsets of observations,
providing local pictures which are then stitched back together to give a global picture.
At first sight,
this differs from our method in that we restrict to subsets of \emph{variables} instead of subsets of observations.
However, if we flip perspectives and think of each genetic variant as an observation,
our method shares common threads, although the ultimate goals and methods for visualization are different.
Future methods for visualization of genomic data
may benefit from other advances in this substantial literature 
\citep[reviewed in][]{vandermaaten2009dimensionality}.

@@

Unlike the other two species,
the method applied separately on all eight chromosomes of \textit{Medicago truncatula} 
showed similar patterns of gradual change in population structure across each chromosome,
with no indications of chromosome-specific patterns.
This consistency suggests that the factor driving the population structure for each chromosome is the same,
as might be caused by varying strengths of linked selection.
To verify that variation in population structure is shared across chromosomes,
we applied the method to all chromosomes together.
Results for chromosome 3 are shown in Figures \ref{fig:mds12_medicago} and \ref{fig:pca_medicago},
and other chromosomes are similar:
% the two types of population structure with higher gene density 
% (colored green and purple in Figure \ref{fig:mds12_medicago})
% are nearly identical across all chromosomes,
% but population structure associated with low gene density (colored orange)
% shows more chromsome-specific patterns (see Supplemental Figure \ref{fig:pca_drosophila_allchr}).
across chromosomes, the high values of the first MDS coordinate coincide with the position of the heterochromatic regions surrounding the centromere,
which often have lower gene density and may therefore be less subject to linked selection \citep{kulikova2001integration,paape2013selection}. 
To verify that this is a possible explanation,
we computed gene densities in each window using gene models in Mt4.0 from \url{jcvi.org} \citep{tang2014improved},
which are shown juxtaposed with
the first MDS coordinate of each window is shown in Figure \ref{fig:mds_medicago},
and are significantly correlated, as shown in Figure \ref{fig:mds_gene_count}.
(Values shown are the number of start and end positions of each predicted mRNA transcript,
divided by two, assigned to the nearest window.)
% that lie closer to SNPs  between the last SNP of each window and the first SNP in the next window.
Other genomic features covary with gene density, however,
so we cannot rule out alternative hypotheses.

@@

There has been substantial debate over the relative impacts of different forms of selection.
These have been difficult to disentangle in part because most concrete theoretical predictions
come in the form of predicted forms of the allele frequency spectrum which are only strictly valid in randomly mating (i.e., unstructured) populations,
and it is unclear to what extent the spatial structure observed in most real populations will affect these predictions.
It is possible that statistics making stronger use of spatial information
will have more luck.

>>>>

In this work we describe variation in population structure using local PCA,
where ``local'' refers to proximity along the genome.
A number of general methods for dimensionality reduction also use a strategy of ``local PCA''
\citep[e.g.,][]{manjon2013diffusion,kambhatla1997dimension,weingessel2000local,roweis2000nonlinear},
performing PCA not on the entire dataset but instead on subsets of observations,
providing local pictures which are then stitched back together to give a global picture.
At first sight,
this differs from our method in that we restrict to subsets of \emph{variables} instead of subsets of observations.
However, if we flip perspectives and think of each genetic variant as an observation,
our method shares common threads, although 
our method does not subsequently use adjacency along the genome,
as we aim to identify similar regions that may be distant.
% Future methods for visualization of genomic data
% may benefit from other advances in this substantial literature 
% \citep[reviewed in][]{vandermaaten2009dimensionality}.

@@

Unlike the other two species,
the method applied separately on all eight chromosomes of \textit{Medicago truncatula} 
showed similar patterns of gradual change in population structure across each chromosome,
with no indications of chromosome-specific patterns.
This consistency suggests that the factor driving the population structure for each chromosome is the same,
as might be caused by varying strengths of linked selection.
To verify that variation in population structure is shared across chromosomes,
we applied the method to all chromosomes together.
Results for chromosome 3 are shown in Figures \ref{fig:mds12_medicago} and \ref{fig:pca_medicago},
and other chromosomes are similar:
% the two types of population structure with higher gene density 
% (colored green and purple in Figure \ref{fig:mds12_medicago})
% are nearly identical across all chromosomes,
% but population structure associated with low gene density (colored orange)
% shows more chromsome-specific patterns (see Supplemental Figure \ref{fig:pca_drosophila_allchr}).
across chromosomes, the high values of the first MDS coordinate coincide with the position of the heterochromatic regions surrounding the centromere,
which often have lower gene density and may therefore be less subject to linked selection.
% Selection on genes in Medicago has been demonstrated by \citet{paape2013selection}. 
To verify that this is a possible explanation,
we computed gene densities in each window using gene models in Mt4.0 from \url{jcvi.org} \citep{tang2014improved},
which are shown juxtaposed with
the first MDS coordinate of each window is shown in Figure \ref{fig:mds_medicago},
and are significantly correlated, as shown in Figure \ref{fig:mds_gene_count}.
(Values shown are the number of start and end positions of each predicted mRNA transcript,
divided by two, assigned to the nearest window.)
% that lie closer to SNPs  between the last SNP of each window and the first SNP in the next window.
Other genomic features covary with gene density, however,
so we cannot rule out alternative hypotheses.

@@

There has been substantial debate over the relative impacts of different forms of selection.
These have been difficult to disentangle in part because most concrete theoretical predictions
come in the form of predicted forms of the allele frequency spectrum which are only strictly valid in randomly mating (i.e., unstructured) populations,
and it is unclear to what extent the spatial structure observed in most real populations will affect these predictions.
It is possible that statistics making stronger use of spatial information
will have more luck.
\plr{expand}
